Data Analytics Introduction Using Python Training

### 2024-11-12 MsaPythonDataAnalytics.qmd
'update_day_4'

@@

Array descriptors

data.shape
data.dtype
data.ndim

@@

@@

## Joins & Merge {visibility="hidden"}

>>>>

a = np.array([1,2,3])
b = np.array([(1.5,2,3), (4,5,6)], dtype = float)
c = np.array([[(1.5,2,3), (4,5,6)],[(3,2,1), (4,5,6)]], dtype = float)

np.zeros((3,4)) #Create an array of zeros
np.ones((2,3,4),dtype=np.int16) #Create an array of ones
d = np.arange(10,25,5)#Create an array of evenly spaced values (step value)
np.linspace(0,2,9) #Create an array of evenlyspaced values (number of samples)
e = np.full((2,2),7)#Create a constant array
f = np.eye(2) #Create a 2X2 identity matrix
np.random.random((2,2)) #Create an array with random values
np.empty((3,2)) #Create an empty array

@@

### Array descriptors & inspectors

data.shape #Array dimensions
len(a)#Length of array
b.ndim #Number of array dimensions
e.size #Number of array elements
b.dtype  #Data type of array elements
b.dtype.name  #Name of data type
b.astype(int). #Convert an array to a different type

@@

### Array Arithmetics

g = a - b. #Subtraction

np.subtract(a,b) #Subtraction
b + a #Addition

np.add(b,a) #Addition 
a/b #Division

np.divide(a,b) #Division 
a * b #Multiplication

np.multiply(a,b) #Multiplication 
np.exp(b) #Exponentiation
np.sqrt(b) #Square root
np.sin(a)  #Print sines of an array
np.cos(b) #Elementwise cosine
np.log(a)#Elementwise natural logarithm
e.dot(f) #Dot product

**Comparison**

a == b #Elementwise comparison

a< 2 #Elementwise comparison

np.array_equal(a, b) #Arraywise comparison
```

**Sorting Arrays**

a.sort() #Sort an array
c.sort(axis=0) #Sort the elements of an array's axis
```

**Subsetting, slicing, indexing

a[2] #Select the element at the 2nd index

b[1,2] #Select the element at row 1 column 2(equivalent to b[1][2])

a[0:2]#Select items at index 0 and 1

b[0:2,1] #Select items at rows 0 and 1 in column 1

b[:1] #Select all items at row0(equivalent to b[0:1, :])

c[1,...] #Same as[1,:,:]

a[ : : -1] #Reversed array a array([3, 2, 1])

a[a<2] #Select elements from a less than 2

b[[1,0,1, 0],[0,1, 2, 0]] #Select elements(1,0),(0,1),(1,2) and(0,0)

b[[1,0,1, 0]][:,[0,1,2,0]] #Select a subset of the matrix’s rows and columns

### Array Manipulation

i = np.transpose(b) #Permute array dimensions
i.T #Permute array dimensions

b.ravel() #Flatten the array
g.reshape(3, -2) #Reshape, but don’t change data

h.resize((2,6)) #Return a new arraywith shape(2,6)
np.append(h,g) #Append items to an array
np.insert(a,1,5)  #Insert items in an array
np.delete(a,[1])  #Delete items from an array

np.concatenate((a,d),axis=0) #Concatenate arrays

np.vstack((a,b) #Stack arrays vertically(row wise)

np.r_[e,f] #Stack arrays vertically(row wise)
np.hstack((e,f)) #Stack arrays horizontally(column wise)

np.column_stack((a,d)) #Create stacked column wise arrays

np.c_[a,d] #Create stacked column wise arrays

np.hsplit(a,3) #Split the array horizontally at the 3rd index

np.vsplit(c,2) #Split the array vertically at the 2nd index

@@

### Joins & Merge

**Merge***

df1 = pd.DataFrame({'employee': ['Bob', 'Jake', 'Lisa', 'Sue'],
                    'group': ['Accounting', 'Engineering', 'Engineering', 'HR']})
df2 = pd.DataFrame({'employee': ['Lisa', 'Bob', 'Jake', 'Sue'],
                    'hire_date': [2004, 2008, 2012, 2014]})

df3 = pd.merge(df1, df2)
df3

df4 = pd.DataFrame({'group': ['Accounting', 'Engineering', 'HR'],
                    'supervisor': ['Carly', 'Guido', 'Steve']})

pd.merge(df3, df4)

#Many-to-many joins

df5 = pd.DataFrame({'group': ['Accounting', 'Accounting',
                              'Engineering', 'Engineering', 'HR', 'HR'],
                    'skills': ['math', 'spreadsheets', 'coding', 'linux',
                               'spreadsheets', 'organization']})

pd.merge(df1, df5)

df3 = pd.DataFrame({'name': ['Bob', 'Jake', 'Lisa', 'Sue'],
                    'salary': [70000, 80000, 120000, 90000]})

pd.merge(df1, df3, left_on="employee", right_on="name")

#Drop redundant column
pd.merge(df1, df3, left_on="employee", right_on="name").drop('name', axis=1)

**Join**

df1a = df1.set_index('employee')
df2a = df2.set_index('employee')

pd.merge(df1a, df2a, left_index=True, right_index=True)

#For convenience, DataFrames implement the join() method, which performs a merge that defaults to joining on indices:

df1a.join(df2a)

df6 = pd.DataFrame({'name': ['Peter', 'Paul', 'Mary'],
                    'food': ['fish', 'beans', 'bread']},
                   columns=['name', 'food'])
df7 = pd.DataFrame({'name': ['Mary', 'Joseph'],
                    'drink': ['wine', 'beer']},
                   columns=['name', 'drink'])

pd.merge(df6, df7)

#Inner Join

pd.merge(df6, df7, how='inner')

#Other options for the how keyword are 'outer', 'left', and 'right'. An outer join returns a join over the union of the input columns, and fills in all missing values with NAs

pd.merge(df6, df7, how='outer')

pd.merge(df6, df7, how='left')

### Let us get our hands dirty

From the google drive link request for access and download the data [google drive](https://drive.google.com/drive/folders/1BkcXX95KMn_6b_Y8KwjAkGjzMtogWxd7?usp=sharing)

1. Download the state data
1. Merge the three different csv's
1. Rank the population density for the different states in 2010
  - population density calculation = population size / state size area
1. Top 3 states
